---
title: "&nbsp;"
author: "@hrbrmstr"
date: March 07, 2018
output: rmarkdown::github_document
---
```{r init, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, dev="png",
                      fig.retina = 2, fig.width = 10, fig.height = 4)
```

I work with internet-scale data and do my fair share of macro-analyses on
vulnerabilities. I use the R [`semver`]() package for most of my work and
wanted to blather on a bit about it since it's super-helpful for this work
and doesn't get the attention it deserves. `semver` makes it possible to
create charts like this:

which are very helpful in when conducting exposure analytics.

We'll need a few packages to help us along the way:

```{r libs}
library(here) # file mgmt
library(semver) # the whole purpose of the blog post
library(rvest) # we'll need this to get version->year mappings
library(stringi) # b/c I'm still too lazy to switch to ore
library(hrbrthemes) # pretty graphs
library(tidyverse) # sane data processing idioms
```

By issuing a `stats` command to a `memcached` instance you can get a full list
of statistics for the server. The recent newsmaking DDoS used this feature in
conjunction with address spoofing to create 30 minutes of chaos for GitHub.

I sent a `stats` command (followed by a newline) to a vanilla `memcached`
installation and it returned 53 lines (1108 bytes) of `STAT` results that look
something like this:

    STAT pid 7646
    STAT uptime 141
    STAT time 1520447469
    STAT version 1.4.25 Ubuntu
    STAT libevent 2.0.21-stable
    ...

The `version` bit is what we're after, but there are plenty of other
variables you could just as easily focus on if you use `memcached` in
any production capacity.

I extracted raw version response data from our most recent scan for
open `memcached` servers on the internet. For ethical reasons, I cannot
blindly share the entire raw data set but hit up research@rapid7.com if
you have a need or desire to work with this data.

Let's read it in:

```{r data}
version_strings <- read_lines(here("data", "versions.txt"))
```

And, now take a look at it:

```{r explore}
set.seed(2018-03-07)

sample(version_strings, 50)
```

It's in decent shape, but it needs some work if we're going to do a
version analysis with it. Let's clean it up a bit:

```{r cleanup}
data_frame(
  string = stri_match_first_regex(version_strings, "STAT version (.*)$")[,2]
) -> versions

count(versions, string, sort = TRUE) %>%
  knitr::kable(format="markdown")
```

Much better! However, we really only need the major parts of the
[semantic version string](https://semver.org/) for a macro view, so
let's remove non-version strings completely and extract just the
_major_, _minor_ and _patch_ bits:

```{r moar_cleanup}
filter(versions, !stri_detect_fixed(string, "UNKNOWN")) %>% # get rid of things we can't use
  mutate(string = stri_match_first_regex(
    string, "([[:digit:]]+\\.[[:digit:]]+\\.[[:digit:]]+)")[,2] # for a macro-view, the discrete sub-versions aren't important
  ) -> versions

count(versions, string, sort = TRUE) %>%
  knitr::kable(format="markdown")
```

Much, much better! Now, let's dig into the versions a bit.

Using `semver` is dirt-simple. Just use `parse_version()` to get
the usable bits out:

```{r semver01
ex_ver <- semver::parse_version(head(versions$string[1]))

ex_ver

str(ex_ver)
```

It's a special class, referncing an external pointer (the package relies on
an underling C++ library and wraps everything up in a bow for us).

These objects can be compared, ordered, sorted, etc but I tend to just
turn the parsed versions into a data frame that can be associated back
with the main strings. That way we keep things pretty tidy and have tons
of flexibility.

```{r}
bind_cols(
  versions,
  pull(versions, string) %>%
    semver::parse_version() %>%
    as.data.frame()
) %>%
  arrange(major, minor, patch) %>%
  mutate(string = factor(string, levels = unique(string))) -> versions

versions
```

Now we have a tidy data frame and I did the extra step of creating an ordered
`factor` out of the version strings since they are ordinal values. With just
this step, we have everything we need to do a basic plot shoing the
version counts in-order:

```{r}
count(versions, string) %>%
  ggplot() +
  geom_segment(
    aes(string, n, xend = string, yend = 0),
    size = 2, color = "lightslategray"
  ) +
  scale_y_comma() +
  labs(
    x = "memcached version", y = "# instances found",
    title = "Distribution of memcached versions"
  ) +
  theme_ipsum_ps(grid = "Y") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

That chart is informative on its own since we get the perspective that there
are some really old versions exposed. But, how old are they? Projects like
Chrome or Firefox churn through versions regularly/quickly (on purpose).
To make more sense out of this we'll need more info on releases.

This is where things can get ugly for folks who do not have commercial
software management databases handy (or are analyzing a piece of software
that hasn't made it to one of those databases yet). The `memcached` project
maintains a [wiki page](https://github.com/memcached/memcached/wiki/ReleaseNotes)
of version history that's mostly complete, and definitely complete enough for
this exercise. It _will_ some processing before we can associate a version
to a year.

GitHub does not allow scraping of their site and --- off the top of my head ---
I do not know if there is a "wiki" API endpoint, but I _do_ know that you can
tack on `.wiki.git` to the end of a GitHub repo to clone the wiki pages, so
we'll use that knowledge and the `git2r` package to gain access to the
`ReleaseNotes.md` file that has the data we need:

```{r}
td <- tempfile("wiki", fileext="git") # temporary "directory"

dir.create(td)

git2r::clone(
  url = "git@github.com:memcached/memcached.wiki.git",
  local_path = td,
  credentials = git2r::cred_ssh_key() # need GH ssh keys setup!
) -> repo

read_lines(file.path(repo@path, "ReleaseNotes.md")) %>%
  keep(stri_detect_fixed, "[[ReleaseNotes") %>%
  stri_replace_first_regex(" \\* \\[\\[.*]] ", "") %>%
  stri_split_fixed(" ", 2, simplify = TRUE) %>%
  as_data_frame() %>%
  set_names(c("string", "release_year")) %>%
  mutate(string = stri_trim_both(string)) %>%
  mutate(release_year = stri_replace_first_fixed(release_year, "(", "")) %>% # remove leading parens
  mutate(release_year = stri_replace_all_regex(release_year, "\\-.*$", "")) %>% # we only want year so remove remaining date bits from easy ones
  mutate(release_year = stri_replace_all_regex(release_year, "^.*, ", "")) %>% # take care of most of the rest of the ugly ones
  mutate(release_year = stri_replace_all_regex(release_year, "^[[:alpha:]].* ", "")) %>% # take care of the straggler
  mutate(release_year = stri_replace_last_fixed(release_year, ")", "")) %>% # remove any trailing parens
  mutate(release_year = as.numeric(release_year)) -> memcached_releases # make it numeric

unlink(td, recursive = TRUE) # cleanup the git repo we downloaded

memcached_releases
```

We have more versions in our internet-scraped `memcached` `versions` data
set than this wiki page has on it, so we need to restrict the official
release history to what we have. Then, we only want a single instance of
each year for the annotations, so we'll have to do some further processing:

```{r}
filter(memcached_releases, string %in% unique(versions$string)) %>%
  mutate(string = factor(string, levels = levels(versions$string))) %>%
  group_by(release_year) %>%
  arrange(desc(string)) %>%
  slice(1) %>%
  ungroup() -> annotation_df

knitr::kable(annotation_df, "markdown")
```

Now, we're ready to add the annotation layers! We'll take a blind stab
at it before adding in further aesthetic customizations:

```{r}
version_counts <- count(versions, string) # no piping this time

ggplot() +
  geom_blank(data = version_counts,aes(string, n)) + # prime the scales
  geom_vline(
    data = annotation_df, aes(xintercept = as.numeric(string)),
    size = 0.5, linetype = "dotted", color = "orange"
  ) +
  geom_segment(
    data = version_counts,
    aes(string, n, xend = string, yend = 0),
    size = 2, color = "lightslategray"
  ) +
  geom_label(
    data = annotation_df, aes(string, Inf, label=release_year),
    family = font_ps, size = 2.5, color = "lightslateblue",
    hjust = 0, vjust = 1, label.size = 0
  ) +
  scale_y_comma() +
  labs(
    x = "memcached version", y = "# instances found",
    title = "Distribution of memcached versions"
  ) +
  theme_ipsum_ps(grid = "Y") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

_Almost_ got it in ggpar 1! We need to tweak this so that the labels
do not overlap each other and do not obstruct the segment bars. We can
do most of this work in `geom_segment()` itself, plus add a bit of a tweal
to the Y axis scale:

```{r}
ggplot() +
  geom_blank(data = version_counts,aes(string, n)) + # prime the scales
  geom_vline(
    data = annotation_df, aes(xintercept = as.numeric(string)),
    size = 0.5, linetype = "dotted", color = "orange"
  ) +
  geom_segment(
    data = version_counts,
    aes(string, n, xend = string, yend = 0),
    size = 2, color = "lightslategray"
  ) +
  geom_label(
    data = annotation_df, aes(string, Inf, label=release_year), vjust = 1,
    family = font_ps, size = 2.5, color = "lightslateblue", label.size = 0,
    hjust = c(1, 0, 1, 1, 0, 1, 0, 0, 1, 0),
    nudge_x = c(-0.1, 0.1, -0.1, -0.1, 0.1, -0.1, 0.1, 0.1, -0.1, 0.1)
  ) +
  scale_y_comma(limits = c(0, 2050)) +
  labs(
    x = "memcached version", y = "# instances found",
    title = "Distribution of memcached versions"
  ) +
  theme_ipsum_ps(grid = "Y") +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.5, angle = 90))
```

Now, we have version and year info to we can get a better idea of the scope
of exposure (and, just how much technical debt many organizations have accrued).

With the ordinal version inforamtion we can also perform other statistical
operations as well. All due to the `semver` package.

You can find this R project over at [GitHub](https://github.com/hrbrmstr/a-version)